### 2020-04-18 Thesis-Chapter2-Background.tex
Finish section on US regs

All these realities--government discomfort with absolute privacy, rapidly increasing availability of strong encryption,
and a blossoming computer industry foundationally reliant on cryptography--came to a head in what has become known as
the first ``crypto war.''
% \ac{the-cw1}.

In 1976, the same year Diffie and Hellmann published their research in public key cryptography, the U.S. Congress passed
the \ac{AECA} and declared that strong cryptography is subject to export controls \cite{kehl_right_2015}. In 1991, the
U.S. Senate introduced, but did not pass, a bill mandating access to plain text contents when authorized by law; in
response, Philip Zimmermann released \ac{PGP}, a public key encryption tool to secure email, in order that strong
cryptography be ``made available to the American public before it became illegal to use'' \cite{zimmermann_1996}. In
1993, the Clinton administration introduced the \ac{clipperchip} \cite{press_1993} with the goal of ``providing the
public with strong cryptographic tools without sacrificing the ability of law enforcement and intelligence agencies to
access unencrypted versions of those communications'' \cite{thompson_2015}. Citing the foundational role of cryptography
in security--and the potential of human rights abuses for compromised privacy--industry and prominent technical leaders
reacted negatively to the initiative \cite{kehl_right_2015} \cite{zimmermann_1996}. When security researcher Matt Blaze
discovered flaws that allowed users to subvert the \ac{clipperchip} mechanisms \cite{blaze_protocol_1994}, the proposal
died.

Despite the failure of the \ac{clipperchip}, the debate over access to strong encryption continued, primarily focused on
various key escrow proposals \cite{thompson_2015}. In 1996, two bills were introduced in the US Congress. In the Senate,
it was S.1726, the Pro-C0DE Act of 1996, to abolish the export controls on encryption software \cite{burns_s1726_1996}.
In the House, it was HR.3011, the SAFE Act of 1996, to explicitly allow arbitrarily strong encryption for all legal
activity \cite{goodlatte_hr3011_1996}. Indeed, the SAFE Act went further, including ``provisions that would have barred
the government from creating a mandatory key escrow system and would also have removed export restrictions on most
generally available software and hardware containing encryption'' \cite{kehl_right_2015}. Zimmermann, the author of PGP,
who had by this time endured an investigation by the Customs Service for publishing his work, testified before the
Senate in favor of the Pro-C0DE Act \cite{zimmermann_1996}. In November of that year, the Clinton administration
released an executive order effectively removing export controls on strong encryption products, along the lines of the
Pro-C0DE Act \cite{clinton_1996}.

From 1996 to 1999, the SAFE Act was proposed several times, had hearings, and gained support \cite{kehl_right_2015}. By
this time, there was ``an overwhelming amount of evidence against moving ahead with any key escrow schemes''
\cite{thompson_2015}, and in 1999 the Clinton administration abruptly changed course, adopting almost all SAFE Act
proposals \cite{kehl_right_2015}. This development marks the end of \ac{the-cw1}.

@@

Two developments paved the way for \ac{the-cw2}. The first was actually a non-development: when the Clinton
administration changed its encryption policy, the U.S. House dropped the SAFE Act \cite{kehl_right_2015}. Also un-passed
were the related Pro-C0DE Act, Encrypted Communications Privacy Act (not to be confused with the \textit{Electronic}
Communications Privacy Act \acs{ECPA} of 1986), and Encryption Protects the Rights of Individuals from Violation and
Abuse in CYberspace (E-PRIVACY) Act \cite{leahy_s376_1997} \cite{ashcroft_s2067_1998}. This legislative failure meant
that rather than being written in the pen of law, encryption policy was written in the pencil of executive order.

@@

That brings us to where the debate stands today. A December 2019 U.S. Senate Judiciary Committee featured the top two
U.S. parties both expressing anger over whether data should be ``beyond the reach of the law'' and expressed threats
like ``get your act together, or we will gladly get your act together for you'' \cite{geller_2019}. In March 2020, the
Republican atop the Senate Judiciary Committee introduced S.3398, the EARN IT Act of 2020, which seeks to establish a
16-chair committee with the power to revoke online platforms' liability protection for user-submitted content if the
committee determines the platform is not doing enough to fight \ac{CSAM}, a legal synonym for
child pornography \cite{graham_s3398_2020}. As illustrated by an in-depth article by the \textit{New York Times}
\cite{keller_internet_2019} that likely triggered the December 2019 hearing, the problem is enormous; however, critics
largely see the EARN IT Act as an attempt to indirectly yet effectively outlaw \ac{E2E} encryption \cite{newman_2020}
\cite{pfefferkorn_2020}.

@@

\newcommand{\lawsstart}{\begin{itemize}}
\newcommand{\law}[4]{ % Year, name, citation, description
    \item #1: \textbf{#2} \cite{#3}

@@

The \ac{USAFREEDOMAct}, introduced by the same U.S. House representative as the \ac{USAPATRIOTAct}, passed the senate
almost two years to the day after \ac{snowden}'s leaks began. It contains many reforms to \ac{FISA} and the use of
\acp{NSL}. Most notably, it ends the interpretation of section 215 of the \ac{USAPATRIOTAct} that enabled mass
collection of telephone metadata.

@@

    % TODO: working here
The \ac{clipperchip} was a voluntary initiative from the White House with the goal of ``providing the public
with strong cryptographic tools without sacrificing the ability of law enforcement and intelligence agencies to access
unencrypted versions of those communications'' \cite{thompson_2015}. Citing the foundational role of cryptography in
security--and the potential of human rights abuses for compromised privacy--industry and prominent technical leaders
reacted negatively to the initiative \cite{kehl_right_2015} \cite{zimmermann_1996}. When security researcher Matt Blaze
discovered flaws that allowed users to subvert the \ac{clipperchip} mechanisms \cite{blaze_protocol_1994}, the proposal
died.

@@

    \law{1996}{The SAFE Act}{goodlatte_hr3011_1996}{

@@

    \law{1999}{Cyberspace Electronic Security Act}{administration_CESA_1999}{

% Bill by the White House

@@

% Bill _on behalf of_ the White House; watered down CESA, mostly CFAA updates

@@

@@

% federal SE protections

@@

    \law{2019}{The ENCRYPT Act}{lieu_hr4170_2019}{

% state SE protections

@@

    \law{2020}{The EARN IT Act}{graham_s3398_2020}{

>>>>

All these realities--government discomfort with absolute privacy, rapidly increasing availability of strong encryption,
and a blossoming computer industry foundationally reliant on cryptography--came to a head in what has become known as
the first ``crypto war.''

In 1976, the same year Diffie and Hellmann published their research in public key cryptography, the U.S. Congress passed
the \ac{AECA} and declared that strong cryptography is subject to export controls \cite{kehl_right_2015}. In 1991, the
U.S. Senate introduced, but did not pass, a bill mandating access to plain text contents when authorized by law; in
response, Philip Zimmermann released \ac{PGP}, a public key encryption tool to secure email, in order that strong
cryptography be ``made available to the American public before it became illegal to use'' \cite{zimmermann_1996}. In
1993, the Clinton administration introduced the \ac{clipperchip} \cite{press_1993} with the goal of ``providing the
public with strong cryptographic tools without sacrificing the ability of law enforcement and intelligence agencies to
access unencrypted versions of those communications'' \cite{thompson_2015}. Citing the foundational role of cryptography
in security--and the potential of human rights abuses for compromised privacy--industry and prominent technical leaders
reacted negatively to the initiative \cite{kehl_right_2015} \cite{zimmermann_1996}. When a prominent security researcher
discovered flaws that allowed users to subvert the \ac{clipperchip} mechanisms \cite{blaze_protocol_1994}, the proposal
died.

Despite the failure of the \ac{clipperchip}, the debate over access to strong encryption continued, primarily focused on
various key escrow proposals \cite{thompson_2015}. In 1996, two bills were introduced in the US Congress. In the Senate,
it was S.1726, the \ac{PROCODEAct} of 1996, to abolish the export controls on encryption software
\cite{burns_s1726_1996}. In the House, it was HR.3011, the \ac{SAFEAct} of 1996, to explicitly allow arbitrarily strong
encryption for all legal activity \cite{goodlatte_hr3011_1996}. Indeed, the \ac{SAFEAct} went further, including
``provisions that would have barred the government from creating a mandatory key escrow system and would also have
removed export restrictions on most generally available software and hardware containing encryption''
\cite{kehl_right_2015}. Zimmermann, the author of PGP, who had by this time endured an investigation by the Customs
Service for publishing his work, testified before the Senate in favor of the \ac{PROCODEAct} \cite{zimmermann_1996}. In
November of that year, the Clinton administration released an executive order effectively removing export controls on
strong encryption products, along the lines of the \ac{PROCODEAct} \cite{clinton_1996}.

From 1996 to 1999, the \ac{SAFEAct} was proposed several times, had hearings, and gained support \cite{kehl_right_2015}.
By this time, there was ``an overwhelming amount of evidence against moving ahead with any key escrow schemes''
\cite{thompson_2015}, and in 1999 the Clinton administration abruptly changed course, adopting almost all \ac{SAFEAct}
proposals \cite{kehl_right_2015}. This development marks the end of \ac{the-cw1}.

@@

Two developments paved the way for \ac{the-cw2}. The first was actually a non-development: when the Clinton
administration changed its encryption policy, the U.S. House dropped the \ac{SAFEAct} \cite{goodlatte_hr3011_1996}. The
\ac{PROCODEAct} and other encryption-related bills went un-passed as well \cite{burns_s1726_1996} \cite{leahy_s376_1997}
\cite{ashcroft_s2067_1998} \cite{goss_hr2616_1999} \cite{goss_hr2617_1999} \cite{administration_CESA_1999}
\cite{leahy_s3083_2000}. This legislative failure meant that rather than being written in the pen of law, encryption
policy was written in the pencil of executive order.

% TODO: add reference to future, comprehensive Appendix of all proposed encryption laws

@@

That brings us to where the debate stands today. A December 2019 U.S. Senate Judiciary Committee featured the top two
U.S. parties both expressing anger over whether data should be ``beyond the reach of the law'' and expressed threats
like ``get your act together, or we will gladly get your act together for you'' \cite{geller_2019}. In March 2020, the
Republican atop the Senate Judiciary Committee introduced S.3398, the \ac{EARNITAct} of 2020, which seeks to establish a
16-chair committee with the power to revoke online platforms' liability protection for user-submitted content if the
committee determines the platform is not doing enough to fight \ac{CSAM}, a legal synonym for child pornography
\cite{graham_s3398_2020}. As illustrated by an in-depth article by the \textit{New York Times}
\cite{keller_internet_2019} that likely triggered the December 2019 hearing, the problem is enormous; however, critics
largely see the \ac{EARNITAct} as an attempt to indirectly yet effectively outlaw \ac{E2EE} \cite{newman_2020}
\cite{pfefferkorn_2020}.

@@

\newcommand{\lawsstart}{\begin{itemize}}
\newcommand{\law}[4]{ % Year, name, citation, description
    \item #1: \textbf{#2} \cite{#3} \nopagebreak

@@

The \ac{USAFREEDOMAct}, introduced by the same U.S. House representative as the \ac{USAPATRIOTAct}, passed the Senate
almost two years to the day after \ac{snowden}'s leaks began. It contains many reforms to \ac{FISA} and the use of
\acp{NSL}. Most notably, it ends the interpretation of section 215 of the \ac{USAPATRIOTAct} that enabled mass
collection of telephone metadata.

@@

As described in \mysec{sec-history-cw1}, the \ac{clipperchip} was a voluntary \ac{EA} initiative from the White House.
The initiative targeted \ac{DIM} and established the Escrowed Encryption Standard, which includes an encrypted copy of
the session key in the message \cite{blaze_protocol_1994}. The initiative received a mixed reception and prompted much
debate \cite{kehl_right_2015}. When a security researcher discovered flaws that allowed users to subvert the
\ac{clipperchip} mechanisms, the proposal died \cite{blaze_protocol_1994}.

@@

    \law{1996}{The \acf{SAFEAct}}{goodlatte_hr3011_1996}{

The \ac{SAFEAct} would have broadly protected ``use any encryption regardless of the encryption algorithm selected,
encryption key length chosen, or implementation technique or medium used,'' excepting ``the unlawful use of encryption
in furtherance of a criminal act.'' It also would have lifted export controls. The bill grew support as it was
reintroduced in consecutive congresses, but was abandoned when the Clinton administration adopted the act's policies in
1999 \cite{kehl_right_2015}.

    \law{1996}{The \acf{PROCODEAct}}{burns_s1726_1996}{

The Senate's \ac{PROCODEAct} was very similar to House's \ac{SAFEAct}, though it focused primarily on removing export
controls, and used slightly weaker language protecting all forms of encryption. Like the \ac{SAFEAct}, it was abandoned
when the Clinton administration adopted the act's policies in 1996 \cite{thompson_2015}.

@@

    \law{1999}{\ac{CESA}}{administration_CESA_1999}{

\Ac{CESA} was a legislative proposal by Clinton administration introduced on same day as encryption policy changes that
led to the abandonment of the \ac{SAFEAct}. Section 102 lays out the government position precisely--encryption is an
important tool for information security, but is also used to ``hide unlawful activity by terrorists, drug traffickers,
child pornographers, and other criminals'' (the \ac{horsemen}), and therefore ``[a]ppropriate means must be available to
fulfill these law enforcement objectives.'' The bill would establish voluntary-participation third-party key escrow with
``recovery agents'' that would provide law enforcement with access to \ac{plaintext}. Access would be compelled through
a variety of mechanisms, including warrantless mechanisms through \ac{FISA}. The legislation never got a legislative
sponsor to be officially considered.

@@

After \ac{CESA} failed, the White House proposed watered-down legislation that this time was successfully sponsored by a
senator. This bill would have amended the \ac{ECPA} and \ac{CFAA}. It would have made relatively modest changes to
government data access requirements. Though it received congressional sponsorship, even the sponsor was critical of
parts of the bill, and it died in committee \cite{senate_2000}.

@@

Though this bill was not officially proposed, it represents one of the first legislative attempts of \ac{the-cw2} to
regulate encryption. This draft was released shortly after the close of the Apple vs. FBI case, and would have compelled
Apple to unlock the phone. Though it would not mandate a technical mechanism, the bill would have mandated that
manufacturers and service providers be able to provide \ac{EA}. This mandate notably goes beyond several regulatory
proposal in \ac{the-cw1}. Previewing a common phrase in recent \ac{EA} arguments \cite{geller_2019}, the bill's authors
both used the phrase ``above the law'' in promotion of the bill.

@@

Proposed in reaction to the Compliance with Court Orders Act and anti-encryption sentiment, this bill would have
provided protections against ``mandating the deployment of vulnerabilities in data security technologies'' at the
federal level. It stopped short of explicitly protecting all forms of encryption.

@@

    \law{2019}{The \ac{ENCRYPTAct}}{lieu_hr4170_2019}{

This bill is designed to ``preempt State data security vulnerability mandates and decryption requirements,'' being a
state-level version of the Secure Data Act. This bill goes further to explicitly disallow \ac{EA} requirements in state
law.

% TODO: as of 4/18/20, this bill is still active. Follow up.

@@

    \law{2020}{The \acf{EARNITAct}}{graham_s3398_2020}{

Introduced by the chair of the Senate Judiciary Committee and one of the outspoken encryption critics during the 2019
Senate hearing on encryption and \ac{EA}, the \ac{EARNITAct}, designed to address \ac{CSAM}, has received attention as a
threat to encryption \cite{newman_2020} \cite{pfefferkorn_2020}. As described in \mysec{sec-history-current}, the bill
would revoke online platforms' liability protection for user-submitted content if is does not doing enough to combat
\ac{CSAM}. Critics fear that meeting the regulatory requirements would preclude \ac{E2EE}, and at least one popular
\ac{E2EE} messaging platform has threatened to pull out of the U.S. market if the bill were to pass \cite{lund_2020}.

% TODO: as of 4/18/20, this bill is still active. Follow up.

### 2020-04-18 Thesis-Glossary.tex
Finish section on US regs

% Start acronyms
\newabbreviation{AECA}{AECA}{Arms Export Control Act}
\newabbreviation{AV}{AV}{anti-virus}
\newabbreviation{AWA}{AWA}{All Writs Act}
\newabbreviation{CALEA}{CALEA}{Communications Assistance for Law Enforcement Act}
\newabbreviation{CFAA}{CFAA}{Computer Fraud and Abuse Act}
\newabbreviation{CP}{CP}{child pornography}
\newabbreviation{CSAM}{CSAM}{child sexual abuse material}
\newabbreviation{CSAV}{CSAV}{computer supported argument visualization}
\newabbreviation{DE}{DE}{device encryption}
\newabbreviation{DFD}{DFD}{data flow diagram}
\newabbreviation{DIM}{DIM}{data in motion}
\newabbreviation{DAR}{DAR}{data at rest}
\newabbreviation{E2E}{E2E}{end to end}
\newabbreviation{E2EE}{E2EE}{end to end encryption}
\newabbreviation{EA}{EA}{exceptional access}
\newabbreviation{ECPA}{ECPA}{Electronic Communications Privacy Act}
\newabbreviation{ED}{ED}{ephemeral data}
\newabbreviation{FBI}{FBI}{Federal Bureau of Investigation}
\newabbreviation{FISA}{FISA}{Foreign Intelligence Surveillance Act}
\newabbreviation{FISAAA}{FISAAA}{FISA Amendments Act}
\newabbreviation{FISC}{FISC}{Foreign Intelligence Surveillance Court}
\newabbreviation{IBIS}{IBIS}{issue-based information system}
\newabbreviation{ITAR}{ITAR}{International Traffic in Arms Regulations}
\newabbreviation{KEK}{KEK}{key encrypting key}
\newabbreviation{LE}{LE}{law enforcement}
\newabbreviation{LEA}{LEA}{law enforcement agency}
\newabbreviation{LEO}{LEO}{law enforcement officer}
\newabbreviation{NED}{NED}{non-ephemeral data}
\newabbreviation{NSA}{NSA}{National Security Agency}
\newabbreviation{NSL}{NSL}{National Security Letter}
\newabbreviation{PAA}{PAA}{Protect America Act}
\newabbreviation{PGP}{PGP}{Pretty Good Privacy}
\newabbreviation{PIN}{PIN}{personal identification number}
\newabbreviation{PKI}{PKI}{public key infrastructure}
\newabbreviation{PW}{PW}{password}
\newabbreviation{SE}{SE}{strong encryption}
\newabbreviation{SSO}{SSO}{single sign-on}
\newabbreviation{TLS}{TLS}{Transport Layer Security}
\newabbreviation{VPN}{VPN}{virtual private network}
% End acronyms

@@

\newabbreviation{USAPATRIOTAct}{USA PATRIOT Act}{Uniting and Strengthening America by Providing Appropriate Tools Required to Intercept and Obstruct Terrorism Act}
\newabbreviation{USAFREEDOMAct}{USA FREEDOM Act}{Uniting and Strengthening America by Fulfilling Rights and Ensuring Effective Discipline Over Monitoring Act}

>>>>

% Start acronyms
\newabbreviation{AECA}{AECA}{Arms Export Control Act}
\newabbreviation{AV}{AV}{anti-virus}
\newabbreviation{AWA}{AWA}{All Writs Act}
\newabbreviation{CALEA}{CALEA}{Communications Assistance for Law Enforcement Act}
\newabbreviation{CESA}{CESA}{Cyberspace Electronic Security Act}
\newabbreviation{CFAA}{CFAA}{Computer Fraud and Abuse Act}
\newabbreviation{CP}{CP}{child pornography}
\newabbreviation{CSAM}{CSAM}{child sexual abuse material}
\newabbreviation{CSAV}{CSAV}{computer supported argument visualization}
\newabbreviation{DE}{DE}{device encryption}
\newabbreviation{DFD}{DFD}{data flow diagram}
\newabbreviation{DIM}{DIM}{data in motion}
\newabbreviation{DAR}{DAR}{data at rest}
\newabbreviation{E2E}{E2E}{end to end}
\newabbreviation{E2EE}{E2EE}{end to end encryption}
\newabbreviation{EA}{EA}{exceptional access}
\newabbreviation{ECPA}{ECPA}{Electronic Communications Privacy Act}
\newabbreviation{ED}{ED}{ephemeral data}
\newabbreviation{FBI}{FBI}{Federal Bureau of Investigation}
\newabbreviation{FISA}{FISA}{Foreign Intelligence Surveillance Act}
\newabbreviation{FISAAA}{FISAAA}{FISA Amendments Act}
\newabbreviation{FISC}{FISC}{Foreign Intelligence Surveillance Court}
\newabbreviation{IBIS}{IBIS}{issue-based information system}
\newabbreviation{ITAR}{ITAR}{International Traffic in Arms Regulations}
\newabbreviation{KEK}{KEK}{key encrypting key}
\newabbreviation{LE}{LE}{law enforcement}
\newabbreviation{LEA}{LEA}{law enforcement agency}
\newabbreviation{LEO}{LEO}{law enforcement officer}
\newabbreviation{NED}{NED}{non-ephemeral data}
\newabbreviation{NSA}{NSA}{National Security Agency}
\newabbreviation{NSL}{NSL}{National Security Letter}
\newabbreviation{PAA}{PAA}{Protect America Act}
\newabbreviation{PGP}{PGP}{Pretty Good Privacy}
\newabbreviation{PIN}{PIN}{personal identification number}
\newabbreviation{PKI}{PKI}{public key infrastructure}
\newabbreviation{PW}{PW}{password}
\newabbreviation{SE}{SE}{strong encryption}
\newabbreviation{SSO}{SSO}{single sign-on}
\newabbreviation{TLS}{TLS}{Transport Layer Security}
\newabbreviation{VPN}{VPN}{virtual private network}
% End acronyms

@@

\newabbreviation{USAPATRIOTAct}{USA PATRIOT Act}{Uniting and Strengthening America by Providing Appropriate Tools Required to Intercept and Obstruct Terrorism Act}
\newabbreviation{USAFREEDOMAct}{USA FREEDOM Act}{Uniting and Strengthening America by Fulfilling Rights and Ensuring Effective Discipline Over Monitoring Act}
\newabbreviation{SAFEAct}{SAFE Act}{Security and Freedom Through Encryption Act}
\newabbreviation{PROCODEAct}{Pro-C0DE Act}{Promotion of Commerce On-Line in the Digital Era Act}
\newabbreviation{ENCRYPTAct}{ENCRYPT Act}{Ensuring National Constitutional Rights for Your Private Telecommunications Act}
\newabbreviation{EARNITAct}{EARN IT Act}{Eliminating Abusive and Rampant Neglect of Interactive Technologies Act}
